Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-01-25 paper/zhang-cook-laa-langrene-menendez.Rmd
a coarse go-through of search_better -> SA etc, writing on polish, widefigure for trace plots

Below we introduce three line search algorithms in tour: `search_better`, `search_better_random`, and `search_geodesic`. The first two are simulated annealing algorithms whilst `search_geodesic` evaluates neighbourhood basis to determine the search direction.

@@

`search_better` is a random search device that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is a randomly generated matrix with the same dimension as $\mathbf{A}_{\text{cur}}$. $\mathbf{A}_{l}$ is then orthogonalised to ensure the orthonormal constraint is fulfilled. When a basis is found with index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the search terminates and outputs the basis for guided tour to construct an interpolation path. The next iteration of search begins after adjusting $\alpha$ by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$. The termination condition is when the maximum number of iteration $l_{\max}$ is reached. The algorithm of `search_better` is summarised in Algorithm \ref{random-search}.  A slightly different cooling scheme has been proposed by @posse1995projection to avoid the search space being reduced too fast. A halving parameter $c$ is introduced and $\alpha$ is only adjusted if the last search takes more than $c$ times to find an accepted basis.

@@

Simulated annealing (`search_better_random`) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process as `search_better` but allows a probabilistic acceptance of a basis with lower index value based on the annealing $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when local maxima are present. The algorithm \ref{simulated-annealing} highlights how simulated annealing differs from random search in the inner loop.

@@

@cook1995grand used a line search algorithm on the space of the projection bases. In `search_geodesic`, the search direction is computed using the most prominent direction that deviating an tiny angle of $\delta$ from the current basis. The step size is chosen by optimising the index value along the geodesic direction over an 90 degree angle from $-\pi/4$ to $\pi/4$ along the search direction chosen. The optima $\mathbf{A}_{**}$ is returned for the current iteration if it meets the percentage improve condition or when $l_{\max}$ is reached.  Algorithm \ref{search-geodesic} summarises the inner loop in geodesic search.

@@

In the optimisation algorithms for projection pursuit, the three main elements to record are 1) projection bases: $\mathbf{A}$, 2) index values: $I$, and 3) State: $S$, which labels the observation with detailed stage in the optimisation. Possible values for `search_better` and `search_better_random` include `random_search`,  `new_basis`, and `interpolation`. `search_geodesic` has a wider variety that includes `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`.

@@

Figure \ref{fig:toy-search} shows the searches of two different optimisers: `search_better` and `search_better_random`. Both optimisers progress quickly at the first few iterations, take longer to find better basis in the later iterations, and finish off when hitting the maximum number of try, 499 and 999, respectively. Also, the target basis found in each iteration of `search_better` always has an increased index value while this is not the case for `search_better_random`. This explains why in this example, `search_better_random` takes longer to find the final basis, but in more complicated scenarios, the feature of probabilistic acceptance allows a more holistic search of the basis space and is more likely to find the global optimum.

```{r toy-search, fig.cap="A comparison of search by two otimisers: \\code{search\\_better} and \\code{search\\_better\\_random} on a six-variable dataset \\code{boa6} with holes index on a 2D problem. Both optimisers finish when the maximum number of try is reached and attain similar index value in the end. \\code{search\\_better\\_random} takes longer to get to the final basis because it allows for probabilistic acceptance of inferior basis, which is a feature useful for more complex problem."}
```

@@

```{r toy-interp, fig.cap = "Trace plots of the bases on the interpolation path for two optimisers with the same configuration as the previous figure. The right path shows that to reach a lower target basis, the interpolation actually first passes some higher basis and this could potentially be information to be incorporated into the optimisers."}
```

@@

```{r toy-pca, fig.height=3, fig.cap = "1D projection on the 5-variable dataset \\code{boa5} with two optimisers: pseudo derivative and simulated annealing. All the bases in pseudo derivative has been flipped positive so that the two paths finish close to each other and to the theoretical best."}
```

@@

For a particular index function, the best algorithm to optimise depends on the character of the index and the data. If the index function is smooth and has a single maximum, all of the three algorithms introduced above can find the maximum.  When multiple optima are present, `search_better` may get stuck at a local maximum.  In the case where the index function is non-smooth, `search_geodesic` may even fail to find the maximum. In this section, examples will be presented to outline how the diagnostic plots can be used to compare the performance of optimisers.

@@

A non-monotonic interpolation path has already been given in Figure \ref{fig:toy-interp}: to get to the target basis from the current one, the interpolator may have passed a basis with even higher index value. This motivates the design of an interruption to check whether higher index value can be reached during the interpolation. If such a basis can be found, the optimiser will start the next iteration from that basis instead of the target one. Figure \ref{fig:interruption} contrasts the trace with and without the interruption for the simulated annealing optimiser: `search_better`. The first two interpolation are identical while in the third interpolation, a better basis has been found than the target one. Rather than starting the next iteration from the "target basis" on Time 62, as shown in the left panel, the interruption, on the right panel, starts the next iteration from the "interpolation basis" on Time 61. The interruption has also detected better basis on the fourth interpolation. With the interruption, the optimiser is able to find the final basis with a higher index value (0.865 vs. 0.879).

```{r interruption, fig.asp=0.5,fig.env = "widefigure", fig.cap = "Trace plot of 2D projection on \\code{boa6} data with holes index optimised by \\code{search\\_better}. The purpose of the plot is to compare the effect of implementing the interruption that avoid fully interpolating to the target basis if a better one can be found during the interpolation. With this implementation, the optimiser is able to find a better final basis."}
```

@@

```{r polish, fig.height= 2.5, fig.cap = "Two-D projection on \\code{boa6} data with holes index optimised by \\code{search\\_geodesic}. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish. "}
```

Once the final basis has been found by an algorithm, one may want to push further to investigate whether there is an even better basis in the close neighbourhood. This motivates the polish search where the final basis is supplied as the start of a new guided tour to search for any local breakthrough.

Similar to `search_better` as a stochastic random search, `search_polish` has a different scheme of reducing the search neighbourhood. In each search-interpolation iteration, `search_better` has a fixed neighbourhood parameter alpha and this alpha is reduced by the cooling parameter only after an iteration finishes. On the contrary, `search_polish` allows alpha to be reduced during each iteration to exploit the search in the neighbourhood. Further, to avoid the case where alpha becomes too small and the further search is meaningless, three more stopping criteria have been added, on top of the original `max.tries` limit. These include:

1) the distance between the candidate basis and the current basis needs to be larger than 1e-3;
2) the percentage change of the index value need to be larger than 1e-5; and
3) the alpha parameter on itself needs to be larger than 0.01

Figure \ref{fig:polish} presents the final projections found before and after applying `search_polish` on `search_geodesic`. Polish search improves the index value from `r get_best(holes_2d_geo)$index_val %>% format(digits = 4)`  to `r get_best(holes_2d_geo_polish)$index_val %>% format(digits = 4)` with reduction of weights on the non-informative variables. In terms of the projected data as in Figure \ref{fig:polish}, polish works to sharpen the edges of each cluster.

@@

<!-- ```{r noisy-better-geo,fig.asp=0.5,fig.cap = "One-D projection on `boa5` data with noisy index `norm_kol` optimised by `search_geodesic` and `search_better`. The grey dashed line represents the index value of the theoretical best basis. `search_geodesic` fails to optimise the noisy index while `search_better` has made reasonable improvements to reach the index value close to the one of the theoretical best basis."} -->
<!-- ``` -->

@@

$$\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]$$
With a more complex index function, it is interesting to understand 1) if all the three optimisers can reach the optimum, 2) when a local optimum is presented, whether the simulated annealing algorithm can escape that local optimum and find the global one.

```{r noisy-better-geo, fig.cap = "1D projection on 5\\-variable dataset \\code{boa5} with \\code{norm\\_kol} index. The trace plot shows the failure of pseudo derivative on non-smooth index. The PCA plot also shows that simulated annealing is searching the area close to the theoretical best while pseudo derivative can't couple with the non-continuity in the index."}
```

@@

```{r kol-result, fig.height = 6, fig.cap="One-D projection of data \\code{boa6} on \\code{norm\\_kol} index,  optimised by \\code{search\\_better} and \\code{search\\_better\\_random}. Two different neighbourhood size of 0.5 and 0.7 is used. The finish point of each path is larger than the start point and optimums are separately labelled with * representing the global optimum and x the local optimum The color of the path is based on whether the optimiser finds the global or local optimum, after polishing."}
```

The next experiment compares two optimisers: `search_better` and `search_better_random` for 20 simulation paths. Two search neighbourhood sizes, `alpha`, of 0.5 and 0.7 are also compared to understand the effect of neighbourhood size where a larger value indicating a wider search. Figure \ref{fig:kol-result} shows 80 (20 * 4) PCA projected interpolation paths faceted by optimiser and alpha. Several results can be drawn through the comparison: 1) With `search_better` and `alpha = 0.5`, despite being the simplest and fastest, the optimiser fails to optimise for three instances where they finish neither near the local nor the global optimum. 2) With the same optimiser but a larger neighbourhood size `alpha = 0.7`, more seeds have found the global optimum. 3) Comparing different optimisers `search_better` and `search_better_random` with the same search size of 0.5, `search_better_random` finishes closer to the theoretical for the ones that find the local optimum. This indicates the optimiser is working harder to examine if there is other optimum in the space but a small search neighbourhood size makes it hard to conduct a holistic search of the whole space. 4) With `search_better_random` and a larger neighbourhood size of 0.7, more seeds have found the global optimum. Also some points are getting closer to the position of the theoretical best, as comparing with `alpha = 0.5`, and this would indicate fewer work for the subsequent polish.

>>>>

Below we introduce three line search algorithms in tour: simulated annealing (SA), simulated annealing with jump out (SAJO), and pseudo derivative (PD).

@@

Simulated annealing (SA) is a random search device that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is a randomly generated matrix with the same dimension as $\mathbf{A}_{\text{cur}}$. $\mathbf{A}_{l}$ is then orthogonalised to ensure the orthonormal constraint is fulfilled. When a basis is found with index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the search terminates and outputs the basis for guided tour to construct an interpolation path. The next iteration of search begins after adjusting $\alpha$ by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$. The termination condition is when the maximum number of iteration $l_{\max}$ is reached. The algorithm of simulated annealing is summarised in Algorithm \ref{random-search}.  A slightly different cooling scheme has been proposed by @posse1995projection to avoid the search space being reduced too fast. A halving parameter $c$ is introduced and $\alpha$ is only adjusted if the last search takes more than $c$ times to find an accepted basis.

@@

Simulated annealing with jump out (SAJO) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process but allows a probabilistic acceptance of a inferior basis with lower index value based on the annealing $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when local maxima are present. The algorithm \ref{simulated-annealing} highlights how SAJO differs from SA in the inner loop.

@@

In pseudo derivative [@cook1995grand] method, the search direction is computed using the most prominent direction that deviating an tiny angle of $\delta$ from the current basis. The step size is chosen by optimising the index value along the geodesic direction over an 90 degree angle from $-\pi/4$ to $\pi/4$ along the search direction chosen. The optima $\mathbf{A}_{**}$ is returned for the current iteration if it meets the percentage improve condition or when $l_{\max}$ is reached.  Algorithm \ref{search-geodesic} summarises the inner loop of the pseudo derivative search.

@@

Three main elements are recorded for the projection pursuit optimisers: 1) projection bases: $\mathbf{A}$, 2) index values: $I$, and 3) State: $S$, which labels the observation with detailed stage in the optimisation. For optimiser SA and SAJO, possible values of the state include `random_search`,  `new_basis`, and `interpolation`. pseudo derivative search has a wider variety of state including `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`.

@@

Figure \ref{fig:toy-search} shows the searches of two different optimisers: SA (left) and SAJO (right). Both optimisers progress quickly at the first few iterations, take longer to find better basis in the later iterations, and finish off when hitting the maximum number of try, 499 and 999, respectively. Also, the target basis found on the left panel always has an increased index value in each iteration while this is not the case for the plot on the right. This explains why in this example, the optimiser SAJO takes longer to find the final basis, but in more complicated scenarios, the feature of probabilistic acceptance allows a more holistic search of the basis space and is more likely to find the global optimum.

```{r toy-search, fig.env = "widefigure", fig.cap="A comparison of search by two otimisers: SA and SAJO on a six-variable dataset \\code{boa6} with holes index on a 2D problem. Both optimisers finish when the maximum number of try is reached and attain similar index value in the end. Simulated annealing with jump out (SAJO) takes longer to get to the final basis because it allows for probabilistic acceptance of inferior basis, which is a feature useful for more complex problem."}
```

@@

```{r toy-interp, fig.env = "widefigure", fig.cap = "Trace plots of the bases on the interpolation path for two optimisers with the same configuration as the previous figure. The right path shows that to reach a lower target basis, the interpolation actually first passes some higher basis and this could potentially be information to be incorporated into the optimisers."}
```

@@

```{r toy-pca, fig.height=3, fig.cap = "1D projection on the 5-variable dataset \\code{boa5} with two optimisers: Simulated Annealing (SA) and Pseudo Derivative (PD). All the bases in pseudo derivative has been flipped positive so that the two paths finish close to each other and to the theoretical best."}
```

@@

For a particular index function, the best algorithm to optimise depends on the character of the index and the data. If the index function is smooth and has a single maximum, all of the three algorithms introduced above can find the maximum.  When multiple optima are present, simulated annealing (SA) may get stuck at a local maximum.  In the case where the index function is non-smooth, pseudo derivative search may even fail to find the maximum. In this section, examples will be presented to outline how the diagnostic plots can be used to compare the performance of optimisers.

@@

A non-monotonic interpolation path has already been given in Figure \ref{fig:toy-interp}: to get to the target basis from the current one, the interpolator may have passed a basis with even higher index value. This motivates the design of an interruption to check whether higher index value can be reached during the interpolation. If such a basis can be found, the optimiser will start the next iteration from that basis instead of the target one. Figure \ref{fig:interruption} contrasts the trace with and without the interruption for the simulated annealing optimiser and it can be observed that while the first two interpolation are identical, a better basis has been found than the target one in the third interpolation. Rather than starting the next iteration from the "target basis" on Time 62, as shown in the left panel, the interruption, on the right panel, starts the next iteration from the "interpolation basis" on Time 61. The interruption has also detected better basis on the fourth interpolation. With the interruption, the optimiser is able to find the final basis with a higher index value (0.865 vs. 0.879).

```{r interruption, fig.env = "widefigure", fig.cap = "Trace plot of 2D projection on \\code{boa6} data with holes index optimised by optimiser SA. The purpose of the plot is to compare the effect of implementing the interruption that avoid fully interpolating to the target basis if a better one can be found during the interpolation. With this implementation, the optimiser is able to find a better final basis."}
```

@@

```{r polish, fig.height= 2.5, fig.cap = "Two-D projection on \\code{boa6} data with holes index optimised by pseudo derivative. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish. "}
```

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to find an even better basis. A polish search takes the final basis as the start of a new guided tour to search for local breakthrough.

The polish algorithm is similar to the simulated annealing one with three main distinctions on 1) the number of candidate basis generated each time in the inner loop; 2) the search neighbourhood, `alpha`, and 3) the termination conditions. In simulated annealing, only one candidate basis $\mathbf{A}_l$ is generated every time while the polish algorithm allows for a specified number of basis to be evaluated for each repetition in the inner loop via an argument `n_sample`. The search neighbourhood in simulated annealing is reduced in the outer loop, that is, the alpha for one iteration of search remains the same, while in polish search, `alpha` is reduced once a set of candidate bases fails to make an improvement. To avoid the case the neighbourhood size becoming too small to generate meaningful search, more termination criteria have been added, apart from the original `max.tries` condition and  these include:

1) the distance between the basis found and the current basis needs to be larger than 1e-3;
2) the percentage change of the index value need to be larger than 1e-5; and
3) the alpha parameter on itself needs to be larger than 0.01

Figure \ref{fig:polish} presents the final projections found before (left) and after(right) applying the polish using the pseudo derivative optimiser. Before polishing, the four clusters are clearly shown while the edges of each cluster is relatively unshaved. The polish search start with this basis and on the right panel, the edges of the clusters are better trimmed.

@@

@@

$$\max \left[F_{P}(u) - F_{y}(u)\right]$$
With a more complex index function, it is interesting to understand 1) if all the three optimisers can reach the optimum, 2) when a local optimum is presented, whether the simulated annealing algorithm can escape that local optimum and find the global one.

```{r noisy-better-geo, fig.env = "widefigure", fig.cap = "1D projection on 5\\-variable dataset \\code{boa5} with \\code{norm\\_kol} index. The trace plot shows the failure of pseudo derivative on non-smooth index. The PCA plot also shows that simulated annealing is searching the area close to the theoretical best while pseudo derivative can't couple with the non-continuity in the index."}
```

@@

```{r kol-result, fig.height = 6, fig.cap="1D projection of data \\code{boa6} on \\code{norm\\_kol} index,  optimised by SA and SAJO. Two different neighbourhood size of 0.5 and 0.7 is used. The finish point of each path is larger than the start point and optimums are separately labelled with * representing the global optimum and x the local optimum The color of the path is based on whether the optimiser finds the global or local optimum, after polishing."}
```

The next experiment compares the two simulated annealing optimisers, SA and SAJO,  for 20 simulation paths. Two search neighbourhood sizes, `alpha`, of 0.5 and 0.7 are also compared to understand the effect of neighbourhood size where a larger value indicating a wider search. Figure \ref{fig:kol-result} shows 80 (20 * 4) PCA projected interpolation paths faceted by optimiser and alpha. Several results can be drawn through the comparison: 1) With optimiser SA and a 0.5 neighbourhood search size, despite being the simplest and fastest, the optimiser fails to optimise for three instances where they finish neither near the local nor the global optimum. 2) With the same optimiser but a larger neighbourhood size of 0.7, more seeds have found the global optimum. 3) Comparing between the optimiser SA and SAJO with the same search size of 0.5, SAJO finishes closer to the theoretical for the ones that find the local optimum. This indicates that it is working harder to examine if there is other optimum in the space but a small search neighbourhood size makes it hard to conduct a holistic search of the whole space. 4) With optimiser SAJO and a larger neighbourhood size of 0.7, more seeds have found the global optimum. Also some finish points are getting closer to the position of the theoretical best, as comparing with the 0.5 neighbourhood size, and this would indicate fewer work for the subsequent polish.
